# Covariate balance check
In this file we conduct a covariate balance check, to see if our randomization resulted in a somewhat balanced dataset (across the different treatment and control groups).
The idea 



## Load required libraries
```{r}
library(data.table)
library(ggplot2)
library(nnet)
library(lmtest)
library(tableone)
```


## Loading our data
```{r}
d <- fread(file.path("..", "data", "processed", "CombinedData_DB.csv"))
```




## Visually inspect the data:
```{r}
ggplot(d, aes(x=treatment, fill=factor(gender))) +
  geom_bar(position="fill") + 
  labs(x="Treatment Group", y="Proportion", fill="Gender") +
  theme_minimal()

# Similarly for heavy_items
ggplot(d, aes(x=treatment, fill=factor(heavy_items))) +
  geom_bar(position="fill") + 
  labs(x="Treatment Group", y="Proportion", fill="Heavy Items") +
  theme_minimal()
```

```{r}
# Distribution of group_size across treatment groups
ggplot(d, aes(x=treatment, y=group_size)) +
  geom_boxplot() +
  labs(x="Treatment Group", y="Group Size") +
  theme_minimal()
```


## Conduct statistical tests to determine significance
### Gender (categorical)
```{r}
#Contingency table
gender_table <- table(d$treatment, d$gender)

chisq_gender <- chisq.test(gender_table)
print("Chi-square test for gender:")
print(chisq_gender)

fisher_gender <- fisher.test(gender_table)
print("Fisher's exact test for gender:")
print(fisher_gender)
```


### Heavy items (binary)
```{r}
heavy_items_table <- table(d$treatment, d$heavy_items)

chisq_heavy_items <- chisq.test(heavy_items_table)
print("Chi-square test for heavy items:")
print(chisq_heavy_items)


fisher_heavy_items <- fisher.test(heavy_items_table)
print("Fisher's exact test for heavy items:")
print(fisher_heavy_items)
```



### Group size (continuous)
```{r}
# ANOVA
anova_group_size <- aov(group_size ~ treatment, data=d)
print("ANOVA for group size:")
summary(anova_group_size)

# Kruskal-Wallis test (non-parametric, does not assume normality)
kruskal_group_size <- kruskal.test(group_size ~ treatment, data=d)
print("Kruskal-Wallis test for group size:")
print(kruskal_group_size)
```




## Fitting a null model:
```{r}
#Firstly, ensure that the treatment is considered a categorical variable - and that the control group is considered the baseline (useful for when we need to compare the two different treatments to the control)
d[, treatment := factor(treatment, levels = c("Control poster", "Health poster", "Time poster" ))]

#The idea here is that we assume no covariate influce the treatment assignment
null_model <- multinom(treatment ~ 1, data=d)
```


## Fitting a full model:
```{r}
full_model <- multinom(treatment ~ gender + group_size + heavy_items, data=d)
```
## Comparing the models
```{r}
#I'm not that familiar with this lrtest package, but found the following, which seems to be exactly what I want: https://api.rpubs.com/tomanderson_34/lrt
lr_test <- lrtest(null_model, full_model)
print("Likelihood Ratio Test:")
print(lr_test)
```


## Additional covariate balance checks
```{r}
# 1. Create a table summarizing covariates by treatment
vars <- c("gender", "group_size", "heavy_items")
# Setting test = TRUE performs appropriate hypothesis tests for differences across groups
# Setting printOptions = list(smd = TRUE) includes standardized mean differences (SMDs)
tab1 <- CreateTableOne(vars = vars, strata = "treatment", data = d, test = TRUE)
print(tab1, printOptions = list(smd = TRUE))
```

```{r}
# List of variables to test
vars <- c("gender", "group_size", "heavy_items")

# Function to perform appropriate tests for each variable
perform_tests <- function(data, vars, strata_var) {
  results <- list()
  
  for (var in vars) {
    if (is.factor(data[[var]]) || is.character(data[[var]]) || length(unique(data[[var]])) <= 5) {
      # Categorical variable: Chi-square or Fisher's Exact Test
      cat("Testing categorical variable:", var, "\n")
      contingency_table <- table(data[[var]], data[[strata_var]])
      if (any(contingency_table < 5)) {
        test <- fisher.test(contingency_table, workspace = 2e8)
        results[[var]] <- list(method = "Fisher's Exact Test", p_value = test$p.value)
      } else {
        test <- chisq.test(contingency_table)
        results[[var]] <- list(method = "Chi-square Test", p_value = test$p.value)
      }
    } else {
      # Continuous variable: ANOVA or Kruskal-Wallis Test
      cat("Testing continuous variable:", var, "\n")
      formula <- as.formula(paste(var, "~", strata_var))
      anova_test <- aov(formula, data = data)
      results[[var]] <- list(method = "ANOVA", p_value = summary(anova_test)[[1]]["Pr(>F)"][1])
      
      # Uncomment below if you want a non-parametric alternative
      # kw_test <- kruskal.test(formula, data = data)
      # results[[var]] <- list(method = "Kruskal-Wallis Test", p_value = kw_test$p.value)
    }
  }
  return(results)
}

# Run the tests
test_results <- perform_tests(d, vars, "treatment")

# Print the results
cat("Test Results:\n")
for (var in names(test_results)) {
  cat(var, ":", test_results[[var]]$method, "- p-value:", test_results[[var]]$p_value, "\n")
}
```





















```{r}
# 2. Separate tests for categorical variables:
# Chi-square test for gender
cat("Chi-square test for Gender vs. Treatment:\n")
gender_table <- table(d$treatment, d$gender)
print(gender_table)
print(chisq.test(gender_table))

# Chi-square test for heavy_items
cat("\nChi-square test for Heavy Items vs. Treatment:\n")
heavy_table <- table(d$treatment, d$heavy_items)
print(heavy_table)
print(chisq.test(heavy_table))
```

```{r}
# 3. Test for the continuous variable (group_size):
cat("\nANOVA for Group Size vs. Treatment:\n")
anova_group_size <- aov(group_size ~ treatment, data = d)
print(summary(anova_group_size))
```




